Generic Properties of D-optimal Designs in Infinite Dimensional Bayesian Linear Inverse Problems

### 2023-04-08 latex/vanishing.tex
Almost done?

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% SECTION Analysis of Optimal Designs --- Vanishing Model Error
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Characterization of D-Optimal Designs with Vanishing Model Error}\label{section:vanishing}
In this section we analyze D-optimal designs when no model error is
present. We first derive a nonlinear eigenvalue problem in $\obs$ for
D-optimal designs, and then we characterize D-optimal designs as ones
that make all eigenvalues equal in the nonlinear eigenvalue
problem. Results are summarized in Theorem \ref{thm:char}.

%% \subsection{The Nonlinear Eigenvalue Problem}\label{subsec:eigenproblem}
The necessary first-order condition for D-optimality
\eqref{eq:conditions}, when $\Sigma= \sigma^2I$ (i.e. when $\modcov =
0$) becomes
\begin{equation}\label{eq:eigenproblem}
  \sigma^{-2}\fwd \postcov \fwd^* \obs^* = \obs^* \Xi,
\end{equation}
with $\Xi$ diagonal. We think of \eqref{eq:eigenproblem} as an
eigenvalue problem for the self-adjoint operator $\sigma^{-2}\fwd
\postcov \fwd^*$, where rows of $\obs$, namely $\meas_j,j=1,\dots, m$,
are eigenvectors. However, $\postcov$ depends on $\obs$ and thus, we
refer to \eqref{eq:eigenproblem} as a \emph{nonlinear} eigenvalue
problem. The following two lemmas allow us to gain better
understanding of the nonlinear eigenvalue problem.

\begin{restatable*}[Two applications of Woodbury Matrix Identity]{lemma}{woodbury}\label{lemma:twice woodbury}
  Assume $\fwd \prcov \fwd^*$ is invertible. Then
  \begin{align*}
    \begin{split}
      \fwd( \prcov^{-1} + \sigma^{-2}  \fwd^* \obs^* \obs \fwd )^{-1} \fwd^* 
      %
      %
      = \left ( (\fwd\prcov\fwd^*)^{-1} + \sigma^{-2}  \obs^* \obs \right )^{-1},
    \end{split}
  \end{align*}  
\end{restatable*}

%% The term $(\fwd \prcov \fwd^*)^{-1}$ is the prior precision in
%% $\hilo$, and $\fwd\prcov\fwd^*$ is invertible, as mentioned in
%% Section \ref{section:prelim}. $(\fwd\prcov\fwd^*)^{-1} + \sigma^{-2}
%% \obs^* \obs$ is the posterior precision in $\hilo$.

%% \subsection{Characterizing D-Optimal Designs}\label{subsec:characterization}
%% %% Denote corresponding eigenvectors $\{\ev_i\}_{i=1}^{\infty}$.
%% Denote eigenvalues of $\fwd \prcov \fwd^*$ by
%% $\{\lambda_i\}_{i=1}^{\infty}$, and note they are positive. Lemma
%% \ref{lemma:sim diag} is crucial for the rest of the analysis:

@@

\begin{proposition}[Eigenvectors of a D-optimal design]\label{prop:same ev}
  Let $\obs$ satisfy the nonlinear eigenvalue problem
  \eqref{eq:eigenproblem}. Then $\obs^*\obs$ has the same eigenvectors
  as $\fwd \prcov \fwd^*$.
\end{proposition}
\begin{proof}
  \begin{align}\label{eq:mod conditions}
    \begin{split}
      \obs^* \Xi &= \sigma^{-2}\fwd \postcov \fwd^* \obs^*  \text{ (by \eqref{eq:eigenproblem})}\\
      %
      %
      %
      &= \sigma^{-2} \fwd( \prcov^{-1} + \sigma^{-2}  \fwd^* \obs^* \obs \fwd )^{-1} \fwd^* \obs^*  \text{ (by \eqref{eq:postcov})} \\
      %
      %
      %
      &= \sigma^{-2} \left ( (\fwd\prcov\fwd^*)^{-1} + \sigma^{-2}  \obs^* \obs \right )^{-1} \obs^* \text{ (by Lemma \ref{lemma:twice woodbury})}.
    \end{split}
  \end{align}

@@

We are almost ready to state and prove Theorem \ref{thm:char} that
characterizes D-optimal designs. We just need one final lemma, whose
proof is delegated to the appendix:
\begin{restatable*}[Unit norm decomposition]{lemma}{free}\label{lemma:free}
  Let $M \in \R^{k \times k}$ symmetric positive definite with $\ttr M
  = m$, $m \geq k$. We can find $\func_j \in \R^k,j=1,\dots,m$
  with $\|\func_j\|=1$ and $A = (\func_1,\dots,\func_m)$ such that
  $AA^t = M$.
\end{restatable*}

@@

  Recall from \eqref{eq:postcov}, that the posterior precision is
  $\postcov^{-1} = \prcov^{-1} + \sigma^{-2}\fwd^*\obs^*\obs\fwd$. The
  first statement in part (5) now follows from Lemma \ref{lemma:twice
    woodbury}, while the second statement follows from parts (1) and
  (4).
\end{proof}

@@

    \end{axis}
  \end{tikzpicture}
  \caption{A comparison of the eigenvalues of the pushforward
    posterior precision $(\fwd\prcov\fwd^*)^{-1}
    + \sigma^{-2}\obs^*\obs$ for a D-optimal design (left) and a
    sub-optimal design (right). Both designs are allowed three
    observations. We assume $\sigma^2=1$ and thus, the blue area has
    accumulated height of $3$ in both plots. A D-optimal design (left)
    increases precision (reduces uncertainty) where it is lowest. A
    sub-optimal design (right) does not.
    %% In this example, an observation is taken partly in the
    %% direction of the third eigenvector. The precision for this
    %% eigenvector is bigger even than the posterior precision of the
    %% D-optimal design on the left.
  }
  \label{fig:optimal vs not}
\end{figure}

@@

Consider an inverse problem with $\lambda_1^{-1} = 0.1, \lambda_2^{-1}
= 0.2$ and $\sigma^2=1$. In \eqref{eq:two designs}, we suggest two
D-optimal designs $\obs_1, \obs_2$ such that $\obs_1^*\obs_1 =
\obs_2^*\obs_2$. The former exhibits clusterization ($\meas_3 =
\meas_5 = \ev_1$), while the latter does not. Since both of these
designs are D-optimal, clusterization does not pose any obstruction to
D-optimality.

\begin{equation}\label{eq:two designs}
  \obs_1 =
  \begin{pmatrix}
    \sqrt{0.275}\ev_1 - \sqrt{0.725}\ev_2 \\
    \sqrt{0.275}\ev_1 + \sqrt{0.725}\ev_2 \\
    \ev_1 \\
    \ev_2 \\
    \ev_1 \\    
  \end{pmatrix},
  \obs_2 =
  \begin{pmatrix}
    \sqrt{\frac{3}{8}}\ev_1 - \sqrt{\frac{5}{8}}\ev_2 \\
    \sqrt{\frac{3}{8}}\ev_1 + \sqrt{\frac{5}{8}}\ev_2 \\
    \sqrt{\frac{2}{5}}\ev_1 - \sqrt{\frac{3}{5}}\ev_2 \\
    \sqrt{\frac{2}{5}}\ev_1 + \sqrt{\frac{3}{5}}\ev_2 \\
    \ev_1 \\    
  \end{pmatrix}
\end{equation}

A computer implementation of Lemma \ref{lemma:free} (see code in
Supplementary material) generates $\obs_1$, which exhibits
clusterization. Modifying the numerical values of the prior does not
break the clusterization. Thus, the preference to clusterization of
Lemma \ref{lemma:free} is robust. The vast majority of randomized
numerical simulations (see code in Supplementary material) also give
rise to D-optimal designs that exhibit clustrization. We conclude that
clusterization is a generic property of the relaxed model presented in
this study.

@@

      \addlegendentry{prior}
      \addlegendentry{$o_1o_2$}
      \addlegendentry{$o_3o_4$}
      \addlegendentry{$o_5$}
      %% \addlegendentry{$o_4$}
      %% \addlegendentry{$o_5$}
      %% \addlegendentry{$o_6$}   
    \end{axis}
  \end{tikzpicture}
  \caption{Clusterization (left, $\obs_1$ in \eqref{eq:two designs})
    and non-clusterization (right, $\obs_2$ in \eqref{eq:two designs})
    in D-optimal designs. Posterior precision per eigenvector is
    plotted for $\obs_1$ and $\obs_2$ from \eqref{eq:two
      designs}. Both designs are identical for all practical matters
    --- their posteriors are equal. The left panel shows a D-optimal
    design with clusterization: $\meas_3 = \meas_5$. The right panel
    shows a D-optimal design without clusterization. The contribution
    to the posterior precision due to $\meas_1$ and $\meas_2$ is
    denoted $\meas_1\meas_2$. These contributions are plotted together
    because $\meas_1$ and $\meas_2$ are not in the direction of any
    eigenvector. Similarly for $\meas_3\meas_4$.}
  \label{fig:clusterization}
\end{figure}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% SECTION Analysis of Optimal Designs --- Vanishing Model Error
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Characterization of D-Optimal Designs with Vanishing Model Error}\label{section:vanishing}
We now turn to analyze D-optimal designs when $\modcov = 0$. Our goal
is to prove Theorem \ref{thm:char} --- a characterization of D-optimal
designs.

%% \subsection{The Nonlinear Eigenvalue Problem}\label{subsec:eigenproblem}
The necessary first-order condition for D-optimality summarized in
Theorem \ref{thm:constrained}, when $\Sigma= \sigma^2I$ (i.e. when
$\modcov = 0$) becomes
\begin{equation}\label{eq:eigenproblem}
  \sigma^{-2}\fwd \postcov \fwd^* \obs^* = \obs^* \Xi,
\end{equation}
with $\Xi$ diagonal. We think of \eqref{eq:eigenproblem} as an
eigenvalue problem for the self-adjoint operator $\sigma^{-2}\fwd
\postcov \fwd^*$, where rows of $\obs$, namely $\meas_j,j=1,\dots, m$,
are eigenvectors. However, $\postcov$ depends on $\obs$ and thus, we
refer to \eqref{eq:eigenproblem} as a \emph{nonlinear} eigenvalue
problem. The following Proposition and Lemma are proved in the
appendix:

\begin{restatable*}[Two applications of Woodbury Matrix Identity]{proposition}{woodbury}\label{prop:twice woodbury}
  Assume $\fwd \prcov \fwd^*$ is invertible. Then
  \begin{align*}
    \begin{split}
      \fwd( \prcov^{-1} + \sigma^{-2}  \fwd^* \obs^* \obs \fwd )^{-1} \fwd^* 
      %
      %
      = \left ( (\fwd\prcov\fwd^*)^{-1} + \sigma^{-2}  \obs^* \obs \right )^{-1},
    \end{split}
  \end{align*}  
\end{restatable*}

@@

\begin{proposition}[Eigenvectors of a D-optimal design]\label{prop:same ev}
  Let $\obs$ satisfy the nonlinear eigenvalue problem
  \eqref{eq:eigenproblem}. Then $\obs^*\obs$ has the same eigenvectors
  as $\fwd \prcov \fwd^*$.
\end{proposition}
\begin{proof}
  \begin{align}\label{eq:mod conditions}
    \begin{split}
      \obs^* \Xi &= \sigma^{-2}\fwd \postcov \fwd^* \obs^*  \text{ (by \eqref{eq:eigenproblem})}\\
      %
      %
      %
      &= \sigma^{-2} \fwd( \prcov^{-1} + \sigma^{-2}  \fwd^* \obs^* \obs \fwd )^{-1} \fwd^* \obs^*  \text{ (by \eqref{eq:postcov})} \\
      %
      %
      %
      &= \sigma^{-2} \left ( (\fwd\prcov\fwd^*)^{-1} + \sigma^{-2}  \obs^* \obs \right )^{-1} \obs^* \text{ (by Proposition \ref{prop:twice woodbury})}.
    \end{split}
  \end{align}

@@

We are almost ready to prove Theorem \ref{thm:char} that characterizes
D-optimal designs. We just need one final lemma, whose proof is
delegated to the appendix:
\begin{restatable*}[Unit norm decomposition]{lemma}{free}\label{lemma:free}
  Let $M \in \R^{k \times k}$ symmetric positive definite with $\ttr M
  = m$, $m \geq k$. We can find $\func_j \in \R^k,j=1,\dots,m$
  with $\|\func_j\|=1$ and $A = (\func_1,\dots,\func_m)$ such that
  $AA^t = M$.
\end{restatable*}

@@

  Recall from \eqref{eq:postcov}, that the posterior precision is
  $\postcov^{-1} = \prcov^{-1} + \sigma^{-2}\fwd^*\obs^*\obs\fwd$. The
  first statement in part (5) now follows from Lemma \ref{prop:twice
    woodbury}, while the second statement follows from parts (1) and
  (4).
\end{proof}

@@

    \end{axis}
  \end{tikzpicture}
  \caption{A comparison of the eigenvalues of the pushforward
    posterior precision $(\fwd\prcov\fwd^*)^{-1} +
    \sigma^{-2}\obs^*\obs$ for a D-optimal design (left) and a
    sub-optimal design (right). Both designs are allowed three
    observations. We assume $\sigma^2=1$ and thus, the blue area has
    accumulated height of $3$ in both plots. The D-optimal design
    (left) increases precision (equivalently, reduces uncertainty)
    where it is lowest. The sub-optimal design (right) does not.
    %% In this example, an observation is taken partly in the
    %% direction of the third eigenvector. The precision for this
    %% eigenvector is bigger even than the posterior precision of the
    %% D-optimal design on the left.
  }
  \label{fig:optimal vs not}
\end{figure}

@@

Consider an inverse problem with $\lambda_1^{-1} = 0.1, \lambda_2^{-1}
= 0.2$ and $\sigma^2=1$. In \eqref{eq:two designs}, we suggest two
D-optimal designs $\obs_1, \obs_2$ such that $\obs_1^*\obs_1 =
\obs_2^*\obs_2$. The former exhibits clusterization ($\meas_3 =
\meas_5 = \ev_1$), while the latter does not. Both of these designs
are D-optimal, hence clusterization poses no obstruction to
D-optimality.

\begin{equation}\label{eq:two designs}
  \obs_1 =
  \begin{pmatrix}
    \sqrt{\frac{11}{40}}\ev_1 - \sqrt{\frac{29}{40}}\ev_2 \\
    \sqrt{\frac{11}{40}}\ev_1 + \sqrt{\frac{29}{40}}\ev_2 \\
    \ev_1 \\
    \ev_2 \\
    \ev_1 \\    
  \end{pmatrix},
  \obs_2 =
  \begin{pmatrix}
    \sqrt{\frac{3}{8}}\ev_1 - \sqrt{\frac{5}{8}}\ev_2 \\
    \sqrt{\frac{3}{8}}\ev_1 + \sqrt{\frac{5}{8}}\ev_2 \\
    \sqrt{\frac{2}{5}}\ev_1 - \sqrt{\frac{3}{5}}\ev_2 \\
    \sqrt{\frac{2}{5}}\ev_1 + \sqrt{\frac{3}{5}}\ev_2 \\
    \ev_1 \\    
  \end{pmatrix}
\end{equation}

A computer implementation of Lemma \ref{lemma:free} (see code in
Supplementary material) generates $\obs_1$ as a solution to the
D-optimal design problem mentioned in the previous
paragraph. Modifying the numerical values of the prior does not break
the clusterization. Thus, the preference to clusterization of Lemma
\ref{lemma:free} is robust. The vast majority of randomized numerical
simulations (see code in Supplementary material) also give rise to
D-optimal designs that exhibit clustrization. We conclude that
clusterization poses no obstruction to D-optimality in the relaxed
model presented in this study. Moreover, sensor clusterization is
generic, as far as one accepts the model presented in this study.

@@

      \addlegendentry{prior}
      \addlegendentry{$o_1o_2$}
      \addlegendentry{$o_3o_4$}
      \addlegendentry{$o_5$}
      %% \addlegendentry{$o_4$}
      %% \addlegendentry{$o_5$}
      %% \addlegendentry{$o_6$}   
    \end{axis}
  \end{tikzpicture}
  \caption{Clusterization (left, $\obs_1$ in \eqref{eq:two designs})
    and non-clusterization (right, $\obs_2$ in \eqref{eq:two designs})
    in D-optimal designs. Posterior precision per eigenvector is
    plotted for $\obs_1$ and $\obs_2$ from \eqref{eq:two
      designs}. Both designs are identical for all practical matters
    --- their posteriors are equal. In the left panel, a D-optimal
    design with clusterization is illustrated: $\meas_3 = \meas_5$. In
    the right panel, a D-optimal design without clusterization is
    illustrated. The contribution to the posterior precision due to
    $\meas_1$ and $\meas_2$ is denoted $\meas_1\meas_2$. These
    contributions are plotted together because $\meas_1$ and $\meas_2$
    are not in the direction of any eigenvector. Similarly for
    $\meas_3\meas_4$.}
  \label{fig:clusterization}
\end{figure}
